Handling String-Number Conversion in String Constraint Solving

### 2019-11-19 draft.tex
wording

Currently, even for the above constraint, which is from a simple program, most solvers already fail to provide a correct answer. Therefore, there is an urgent need for developing efficient techniques for solving such combinations of constraints. To make matters worse, JavaScript, which powers most interactive content on the Web and increasingly server-side code with Node.js, has conversions between strings and integers embedded in the core of its semantics.  Other scripting languages do too, but we focus on JavaScript due to its prominence.  To see how string-integer conversion pervades semantics, consider the following program:

@@

A casual glance at the above code reveal no use of strings at all, but the semantics of field access is somewhat unusual in JavaScript: the arrays are indexed by strings, and numeric indexes are converted to strings. This conversion is mandated explicitly by JavaScript semantics: the 2019 edition of ECMAScript \cite{?} requires that $ToPropertyKey$ be called on the element expression (\S{12.3.2.1}), and $ToPropertyKey$ calls {\tt{ToString}} on that value in all but special cases (\S{7.1.14}). Any faithful symbolic execution of JavaScript must handle such conversions for even basic array operations to work correctly; consider the following code snippet that manipulates an array \texttt{x}, with its value shown on the right:

@@

The Line 4 of the code updates the length of \texttt{x} to $13$ and hence the \texttt{resetArray(x)} will update \texttt{x} to an array with $13$ zeros.
Now suppose that the Line 4 of the code above is replace to {\tt{x[k] = 5}}, where $\mathtt{k}$ is an user input string and we want to find an input such that $\mathtt{x[15]=0}$. In this case, one would need to know the value of $\sti{k}$ and hence whether it will affect the size of the array {\tt{x}}. More precisely, in the corresponding string constraint encoding, we use a shadow variable $xLen$ to remember the length of $x$ and update its value after all array assignment. For an assignment $\mathtt{x[k] = v}$, we update the array value and length using the following string constraint (combined the $select$ function from SMT array theory): $x_{i+1} = store(x_i,k, v)$ and $xLen_{i+1} = ite( xLen_{i}> \sti{k}, xLen_{i}, \sti{k})$.
This is a rather straightforward array operation, and so  having only limited capability for handling str/num conversion. will cripple any model of non-trivial JavaScript code.

@@

Now observe that {\tt{resetArray}} assigns to the array in a loop of which the size cannot, in general, be known a priori; as such, falls into the case where neither the string nor the integer in a conversion is constant and hence most SMT solvers will not be able to handle it. More specifically, in some program the assignment {\tt{x[k] = 5}} can happen before calling {\tt{resetArray(x)}}, where {\tt{k}} is an user input string. One would need to know the value of $\sti{k}$ and hence whether it will affect the size of the array {\tt{x}}. This is a rather straightforward array operation, and so this limit will cripple any model of non-trivial JavaScript code. }

@@

Solving string constraint with string-number conversion is a very challenging problem. From the theoretical point of view, this problem is already proven to be undecidable~\cite{day2018satisfiability}. In practice, in our experiments (Section \ref{section:evaluation}), the state-of-the-art string constraint solvers provide only limited support to string-number conversion.

In this paper, we propose a framework that handles efficiently string constraints with string-number conversion. Due to the high complexity of solving the exact problem, our framework uses both over and under-approximation of the string constraints. The over-approximation is for proving UNSAT while the under-approximation is for proving SAT of the string constraint. Both over and under-approximation  fall in a decidable fragment of string constraints that can we can efficiently solve.

@@

Our decision procedure proceeds in two steps: The first step consists in over-approximating the set of input constraints into a one that  falls in the chain-free fragment~\cite{abdulla2019chain}, which is decidable. Observe that we can also over-approximate the input constraint to one that falls in the decidable fragment the acyclic fragment~\cite{abdulla2014string} or the straight-line fragment~\cite{chen2019decision}. Our choice of the   chain-free fragment~\cite{abdulla2019chain} is motivated by the fact that  the chain-free fragment is the {\em largest} known decidable fragment for that class of string constraints as far as we know.  In our settings, we over-approximate the formula $\Phi$ by converting $xy=yx$ to two formulas $\{x_1=xy, x_2=yx\}$ using two auxiliary variables $x_1,x_2$ and replace the constraint $n=\sti{x}$ with $n=-1 \vee (n\neq -1 \wedge x\in [0-9]^*$).  The over-approximation module uses also  some heuristics  to derive the constant value of any side of the constraint $n=\sti{x}$ to refine the over-approximation. For instance, assume in some other cases, we can derive the value of $n=12$ from some integer constraints. In such case, we can further derive the value of $x$ is in the regular language $(0^*12)$. 
Observe that if the over-approximated module returns UNSAT then  our decision procedure   declares that the original formula is also UNSAT and terminates. Since $\Phi$ is satisfiable, the over-approximation module will not return UNSAT.

@@

The second step of our decision procedure is only enabled  if the first step returns SAT. In this case,  our decision procedure  restricts the domain of each string variables  to strings that obey some predefined and parameterized pattern. By adjusting the parameters, one can easily enlarge or prune out the potential solution space. In this paper, we propose to use patterns defined by \emph{parametric flat automata} (PFA). A PFA is a {\em flat} finite state automaton consisting of a predefined sequence of loops each of fixed length (see Figure \ref{fig:sfa_def}).  The approach based on PFA is very flexible yet allows very efficiently manipulation. One can pick PFA of \emph{arbitrary structure} for the search space restriction and reduce the string constraint solving problem to a linear formula satisfiability problem in polynomial-time.  Furthermore, we associate for each character a unique natural number. This allows us to assume that the input alphabet is a subset of natural numbers. To avoids the \textit{alphabet explosion problem} that the approach in~\cite{abdulla2017flatten} suffers,  we  labeled each transition between two states of a  PFA by a unique  \emph{character} variable (whose domain is the set of natural numbers) instead of having a transition between each two states for each symbol in the alphabet.    
Such reduction based on PFA is the key to handle string-number conversion efficiently. When we restrict the solution space to those defined by PFAs, the original undecidable problem (i.e., string constraint with string-number conversion) is reduced to the satisfiability problem of linear formulas that can  be solved efficiently.

@@

In the following,  we try to explain the construction of the linear formula using the formula $\Phi$ as an example. Assume that we project the domain of $x$ and $y$ to the PFA in Figure~\ref{fig:sfa} (a) and (b), respectively. The variables $v_1$, $v_2$, $v_3$, $v_4$ in the figure are \emph{character} variables.Thus,  $v_1$, $v_2$, $v_3$, $v_4$ are also integer variables. For example, from the formula $ y \in (12)^+$, we may derive $v_3=1$ and $v_4=2$.

The linear formula produced after the domain restriction will be over variables $v_1$, $v_2$, $v_3$, $v_4$,  as well as, the number of occurrences of each character variables $\#v_1$, $\#v_2$, $\#v_3$, $\#v_4$. Each model of the linear formula encodes a model of the string constraint. For example, $x=``12''$ and $y=``1212''$ is encoded by the assignment $(v_1,v_2,v_3,v_4,\#v_1,\#v_2,\#v_3,\#v_4)= (1,2,1,2,1,1,2,2)$. The assignment says, for example, that $x$ is the parametric word obtained by traversing the loop of $A_x$ once (because $\#v_1 = \#v_2 = 1$), which is $v_1v_2$. Under the assignment $v_1=1$ and $v_2=2$, we obtain $x=``12''$.

>>>>

Currently, even for the above constraint, which is from a simple program, most solvers fail to provide a correct answer. Therefore, there is an urgent need for developing efficient techniques for solving such combinations of constraints. To make matters worse, JavaScript, which powers most interactive content on the Web and increasingly server-side code with Node.js, has conversions between strings and integers embedded in the core of its semantics.  Other scripting languages do too, but we focus on JavaScript due to its prominence.  To see how string-integer conversion pervades semantics, consider the following program:

@@

A casual glance at the above code reveals no use of strings at all, but the semantics of field access is somewhat unusual in JavaScript: the arrays are indexed by strings, and numeric indexes are converted to strings. This conversion is mandated explicitly by JavaScript semantics: the 2019 edition of ECMAScript \cite{?} requires that $ToPropertyKey$ be called on the element expression (\S{12.3.2.1}), and $ToPropertyKey$ calls {\tt{ToString}} on that value in all but special cases (\S{7.1.14}). Any faithful symbolic execution of JavaScript must handle such conversions for even basic array operations to work correctly; consider the following code snippet that manipulates an array \texttt{x}, with its value shown on the right:

@@

The Line 4 of the code updates the length of \texttt{x} to $13$ and hence the \texttt{resetArray(x)} will update \texttt{x} to an array with $13$ zeros.
Now suppose that the Line 4 of the code above becomes {\tt{x[k] = 5}}, where $\mathtt{k}$ is an user input string and we want to find an input such that $\mathtt{x[15]=0}$. In this case, one would need to know the value of $\sti{k}$ and hence whether it will affect the size of the array {\tt{x}}. More precisely, in the corresponding string constraint encoding, we use a shadow variable $xLen$ to remember the length of $x$ and update its value after all array assignment. For an assignment $\mathtt{x[k] = v}$, we update the array value and length using the following string constraint (combined the $select$ function from SMT array theory): $x_{i+1} = store(x_i,k, v)$ and $xLen_{i+1} = ite( xLen_{i}> \sti{k}, xLen_{i}, \sti{k})$.
This is a rather straightforward array operation, and so  having only limited capability for handling str/num conversion. will cripple any model of non-trivial JavaScript code.

@@

Now observe that {\tt{resetArray}} assigns to the array in a loop of which the size cannot, in general, be known a priori; as such, it falls into the case where neither the string nor the integer in a conversion is constant and hence most SMT solvers will not be able to handle it. More specifically, in some program the assignment {\tt{x[k] = 5}} can happen before calling {\tt{resetArray(x)}}, where {\tt{k}} is an user input string. One would need to know the value of $\sti{k}$ and hence whether it will affect the size of the array {\tt{x}}. This is a rather straightforward array operation, and so this limit will cripple any model of non-trivial JavaScript code. }

@@

Solving string constraint with string-number conversion has been proven undecidable~\cite{day2018satisfiability}, and, in practice, in our experiments (Section \ref{section:evaluation}), the state-of-the-art string constraint solvers provide only limited support to string-number conversion.

In this paper, we propose a framework that efficiently handles string constraints with string-number conversion. Due to the diffifulty of solving the problem exactly, our framework combines over and under-approximation of string constraints. The over-approximation is for proving UNSAT when possible, while the under-approximation is for proving SAT when possible. Both over and under-approximation  fall in a decidable fragment of string constraints that can we can efficiently solve.

@@

Our decision procedure has two steps: the first step is over-approximating the set of input constraints into a one that falls in the chain-free fragment~\cite{abdulla2019chain}, which is decidable. Observe that we can also over-approximate the input constraint to one that falls in the decidable fragment the acyclic fragment~\cite{abdulla2014string} or the straight-line fragment~\cite{chen2019decision}. Our choice of the   chain-free fragment~\cite{abdulla2019chain} is motivated by the fact that  the chain-free fragment is the {\em largest} known decidable fragment for that class of string constraints.  In our settings, we over-approximate the formula $\Phi$ by converting $xy=yx$ to two formulas $\{x_1=xy, x_2=yx\}$ using two auxiliary variables $x_1,x_2$ and replace the constraint $n=\sti{x}$ with $n=-1 \vee (n\neq -1 \wedge x\in [0-9]^*$).  The over-approximation module uses also  some heuristics  to derive the constant value of any side of the constraint $n=\sti{x}$ to refine the over-approximation. For instance, assume in some other cases, we can derive the value of $n=12$ from some integer constraints. In such case, we can further derive the value of $x$ is in the regular language $(0^*12)$. 
Observe that if the over-approximated module returns UNSAT then  our decision procedure   declares that the original formula is also UNSAT and terminates. Since $\Phi$ is satisfiable, the over-approximation module will not return UNSAT.

@@

The second step of our decision procedure is only enabled  if the first step returns SAT. In this case,  our decision procedure  restricts the domain of each string variables  to strings that obey some predefined and parameterized pattern. By adjusting the parameters, one can easily enlarge or prune out the potential solution space. In this paper, we propose to use patterns defined by \emph{parametric flat automata} (PFA). A PFA is a {\em flat} finite state automaton consisting of a predefined sequence of loops each of fixed length (see Figure \ref{fig:sfa_def}).  The approach based on PFA is very flexible yet allows very efficient manipulation. One can pick PFA of \emph{arbitrary structure} for search space restriction and reduce the string constraint solving problem to a linear formula satisfiability problem in polynomial-time.  Furthermore, we associate a unique natural number  with each character. This allows us to assume that the input alphabet is a subset of natural numbers. To avoids the \textit{alphabet explosion problem} from which the approach in~\cite{abdulla2017flatten} suffers,  we  label each transition between two states of a  PFA with a unique  \emph{character} variable (whose domain is the set of natural numbers) instead of having a transition between each two states for each symbol in the alphabet.    
Such reduction based on PFA is the key to handle string-number conversion efficiently. When we restrict the solution space to those defined by PFAs, the original undecidable problem (i.e., string constraint with string-number conversion) is reduced to the satisfiability problem of linear formulas that can  be solved efficiently.

@@

In the following,  we explain the construction of the linear formula using the formula $\Phi$ as an example. Assume that we project the domains of $x$ and $y$ to the PFA in Figure~\ref{fig:sfa} (a) and (b), respectively. The variables $v_1$, $v_2$, $v_3$, $v_4$ in the figure are \emph{character} variables. Thus,  $v_1$, $v_2$, $v_3$, $v_4$ are also integer variables. For example, from the formula $ y \in (12)^+$, we may derive $v_3=1$ and $v_4=2$.

The linear formula produced after the domain restriction will be over variables $v_1$, $v_2$, $v_3$, $v_4$,  as well as the number of occurrences of each character variables $\#v_1$, $\#v_2$, $\#v_3$, $\#v_4$.  Each model of the linear formula encodes a model of the string constraint. For example, $x=``12''$ and $y=``1212''$ is encoded by the assignment $(v_1,v_2,v_3,v_4,\#v_1,\#v_2,\#v_3,\#v_4)= (1,2,1,2,1,1,2,2)$. The assignment says, for example, that $x$ is the parametric word obtained by traversing the loop of $A_x$ once (because $\#v_1 = \#v_2 = 1$), which is $v_1v_2$. Under the assignment $v_1=1$ and $v_2=2$, we obtain $x=``12''$.
